## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10
 
if (a <= 5) {
  print(a)
} else {
  print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 18
 
if (a <= 5) {
  print("small")
} else if (a >5 & a <=10) {
  print("medium")
} else if (a > 10 | a== 17) {
  print("large")
} else {
  print("huge")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a/2 != 0) {
  if(a==5){
    print("a is 5")
  }
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) {
  # Paste is a concatenation function
  print(paste(value, "is greater than", value -1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

# change the first 3rd and 6th elements to value 6
for (i in c(1,3,6)){
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while(b < 10) {
  # increase b by 2
  b <- b + 2
  # print b
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
# Start w/ a vector of temps
test_temp <- c(16.5, 26.9, 32.6, 8.6, 28, 28.5, 21,0,21.2, 21, 18.8)

# then we want a func that converts from C to F
c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) +32

  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,7,2), nrow <-2)
test_mat
#1 is for rows, 2 is for columns
apply(test_mat, 1, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.5, 26.9, 32.6, 8.6, 28, 28.5, 21,0,21.2, 21, 18.8), nrow <- 2)

apply(test_temp_mat,1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #1 find the distance between pillars with a width and distance: pillars = function(num_of_pillars, distance, width) {
      pillar <- function(num_of_pillars, distance, width) { 
      if (num_of_pillars < 2)
      {distance = 0}
      else{
      distance <- (num_of_pillars-2) * width + (num_of_pillars-1) * distance * 100
        }
      return(distance)
      }

    #2 find the extra perfect numbers, where the first and last bits are set bits
      extra_perfect <- function(n){
      
      return(seq(1, n, by =2))
      }
      
    #3 remove the first instance of the minimum value
      remove_smallest <- function(numbers){
      # if the length is 0 there are no exciting exhibits
      if (length(numbers) == 0) 
      {
        return(integer(0))
               } 
        #Create copy of number list
        excexh <- numbers
        #find the minimum number for the set
        minValue <- min(excexh)
        #remove minimum number (must work by indice to not delete all instances)
        first <- which(excexh == minValue)[1]
        excexh <- excexh[-first]

      return(excexh)
      }
    #4 list the 2 highest ages from a list, low to high: 
      two_oldest_ages <- function(ages){
       oldest <- max(ages)
      firstI <- which(ages == oldest)[1]
      ages <- ages[-firstI]
      oldest <- c(max(ages), oldest)
      return(oldest)
      }
      
    #5 Find the missing value in the series:
      find_missing <- function(sequence){
      # Your code here, note in R the last item in the sequence needs to use length not negative values
      if( abs(sequence[2] - sequence[1]) <= abs(sequence[length(sequence)] - sequence[length(sequence)-1])) {
        difference <- sequence[2] - sequence[1]
        for (i in 2:length(sequence)) {
          if (sequence[i] - sequence[i-1] != difference) {
            return(sequence[i-1] + difference)
          }
        }
      }
      else {
        difference <- sequence[length(sequence)] - sequence[length(sequence)-1]
          return(sequence[1] + difference)
      }
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    fizzbuzz_definer <- function(integ) {
      if (integ %% 15 == 0) {
        return("fizzbuzz")
      }
      else if (integ %% 5 == 0) {
        return("buzz")
      }
      else if (integ %% 3 == 0) {
        return("fizz")
      }
      #else {    #we could use this section to set the null values and place them in a more horizontal structure
       # return("")
     # }
    }

    test_vector <- c(1:20)
    data_series <- sapply(test_vector, fizzbuzz_definer)
    data_series
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    mariostairs <- function(stairs) {
      s <- 0
      stair = "X"
      while (s < stairs){
        s <- s+1
        #I tried simplly multiplying the 2 values, but got an error.  
        line <- paste(rep(stair, times=  s), collapse = "")
        #Remember \n, needs to be in quotes
        cat(line, "\n")
      
      }
    }

    mariostairs(3)

    ```
